Jovian enables direct inference of germline haplotypes from short reads via generative modeling

### 2023-11-21 manuscript/manuscript.tex
Add jitter to pregen

\title{Jovian enables direct inference of germline haplotypes from short reads via sequence-to-sequence modeling}
\author[1]{Brendan O'Fallon}
\author[1]{Ashini Bolia}
\author[1]{Jacob Durtschi}
\author[1]{Luobin Yang}
\author[1]{Eric Frederickson}
\author[1]{Katherine Noble}
\author[1]{Joshua Coleman}
\author[1]{Hunter Best}

@@

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential component of modern genomics. Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants. Here we describe a new approach that replaces these handcrafted statistical methods with a single, end-to-end deep learning model that directly infers germline haplotypes from short read pileups. Our model, called Jovian, frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method learns to realign complex and ambiguous read mappings to produce accurate haplotype predictions, infers variant genotype and phase accurately, and leverages the local read context to inform predictions about a given position. We also demonstrate that a 2-dimensional positional encoding significantly improved precision of the detected variants. Compared to other callers, sensitivity and precision are higher than GATK HaplotypeCaller, but lower than DeepVariant and Strelka2. 
\end{abstract}

@@

Inference of the DNA sequence variants in a sample is one of the most common and important tasks in modern genomics. Tools that aim to identify variants must overcome several challenges to produce an accurate set of variant calls. For instance, the sequenced reads themselves may contain errors in the form of miscalled bases or PCR-induced repeat length errors, or the tools used to align reads to the reference genome may produce incorrect mappings (e.g. Li et al. 2014, Goldfeder et al. 2016).  In addition, large or complex sequence variants may be obscured by limitations in the read alignment algorithm, and require a sophisticated assembly of reads to identify. Reconstructing real variation and discriminating it from technical error is challenging, and modern variant detection tools employ a variety of statistical and heuristic techniques to achieve high precision and sensitivity (e.g. De Pristo et al. 2011, Nielson et al. 2011).

@@

Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single learned model. We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes. The sequence-to-sequence translation is learned by a transformer-based deep neural network (Vaswani et al. 2017). This approach does not require any of the statistical machinery and heuristic thresholds utilized by most modern callers, and instead uses a deep learning model to reconstruct complete haplotypes directly from aligned NGS reads.

@@

Our model consists of a transformer-based encoder and a simple, fixed-length decoder that produces two output sequences, one for each predicted haplotype. The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 (Vaswani et al. 2017, Paszke et al. 2019). Input tokens are the collection of bases that align to a given reference position, with some modifications described below. We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here. The embedded basecalls are `flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).  Instead of the typical 1-dimensional positional encoding, we employ a 2-dimensional encoding, allowing the model to differentiate across both the token (position) and read dimensions. The 2D encoding implementation follows Wang \& Jyh-Charn (2021).

The decoder consists of a single fully-connected layer with ReLU activation followed by two additional fully-connected components that each generate a single haplotype. The decoder is not autoregressive and does not utilize special start-of-sequence or end-of-sequence tokens. Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs. The decoder is of fixed length and produces two haplotype sequences equal in length to the number of input tokens $g$. The full model architecture is shown in Figure \ref{fig:architecture}.

@@

Training data was obtained from 17 whole genomes sequenced from 5 Genome-in-a-Bottle cell lines. Eight samples were prepared with Illumina Nextera DNA Flex kit and the remainder were prepared with the Illumina TruSeq PCR-free kit. All samples were sequenced on an Illumina NovaSeq 6000 instrument in 2x150 mode, to an approximate read depth of 50. After conversion to fastq, the sequenced reads were aligned to human reference genome GRCh37 with the GEM-mapper (v3, Marco-Sola et al. 2012) and were sorted and converted to CRAM format with samtools version 1.9 (Li et al. 2009). No additional refinements, such as duplicate read marking, base-quality score recalibration, or indel realignment were performed.

@@

Models were trained for 50 epochs unless noted otherwise, using the Adam optimizer, a learning rate of $5e^{-5}$, and batch size of 512. Models were implemented in PyTorch (v1.10, Paszke et al 2019).

>>>>

\title{Jovian enables direct inference of germline haplotypes from short reads via generative modeling}
\author[1]{Brendan O'Fallon}
\author[1]{Ashini Bolia}
\author[1]{Jacob Durtschi}
\author[1]{Luobin Yang}
\author[1]{Eric Frederickson}
\author[1]{Katherine Noble}
\author[1]{Joshua Coleman}
\author[1]{Hunter Best}

@@

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an fundamenal component of modern genomics. Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants. Here we describe a new approach that replaces these handcrafted statistical methods with a single, end-to-end deep learning model that directly infers germline haplotypes from short read pileups. Our model, called Jovian, frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method learns to realign complex and ambiguous read mappings to produce accurate haplotype predictions, infers variant genotype and phase accurately, and leverages the local read context to inform predictions about a given position. We also demonstrate that a 2-dimensional positional encoding significantly improved precision of the detected variants. Compared to other callers, sensitivity and precision are higher than GATK HaplotypeCaller, but lower than DeepVariant and Strelka2. 
\end{abstract}

@@

Inference of the DNA sequence variants in a sample is one of the most common and important tasks in modern genomics. Tools that aim to identify variants must overcome several challenges to produce an accurate set of variant calls. For instance, the sequenced reads themselves may contain errors in the form of miscalled bases or PCR-induced repeat length errors, or the tools used to align reads to the reference genome may produce incorrect mappings (e.g. Li et al. 2014, Goldfeder et al. 2016).  In addition, large or complex sequence variants may be obscured by limitations in the read alignment algorithm, and require a sophisticated reassembly of reads to identify. Reconstructing real variation and discriminating it from technical error is challenging, and modern variant detection tools employ a variety of statistical and heuristic techniques to achieve high precision and sensitivity (e.g. De Pristo et al. 2011, Nielson et al. 2011).

@@

Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single generative model that is learned end-to-end. We recast variant detection as a generative modeling problem in which the outputs are the two haplotype sequences present in the sample. From these generated haplotypes we recover accurate variant calls. Overall, our model architecture resembles that of modern Large Language Models (LLMs). LLMs typically take as input a series of word fragments (tokens) and iteratively create new output tokens, where the model predicts each output token using the list of input tokens as well as the output tokens already generated. In contrast, our model uses columns of aligned bases as input tokens, and predicts two output sequences, one for each diploid haplotype present in the sample. The sequence-to-sequence translation is learned by a transformer-based deep neural network (Vaswani et al. 2017). This approach does not require any of the statistical machinery and heuristic thresholds utilized by most modern callers, and instead uses a deep learning model to reconstruct complete haplotypes directly from aligned NGS reads.

@@

Our model consists of a transformer-based encoder and dual, parallel decoders that each produce a single output haplotype sequence.  The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 2.10 (Vaswani et al. 2017, Paszke et al. 2019). Input tokens are the collection of bases that align to a given reference position, with some modifications described below. We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here. The embedded basecalls are `flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).  Instead of the typical 1-dimensional positional encoding we employ a 2-dimensional encoding, allowing the model to differentiate across both the token (position) and read dimensions. The 2D encoding implementation follows Wang \& Jyh-Charn (2021).

The two decoders each consist of stacked transformer decoder layers with GeLU activations. Each decoder generates a k-mer of length 4 on each prediction step. Because there are $4^4=256$ possible 4-mers the decoders have dimension 256, which is not directly compatible with the encoder dimension, so we add an additional linear layer between the encoder outputs and the decoder layers to convert the model dimensions accordingly.  The full model architecture is shown in Figure \ref{fig:architecture}.

@@

Training data was obtained from 28 whole genomes sequenced from 7 Genome-in-a-Bottle cell lines. Eight samples were prepared with Illumina Nextera DNA Flex kit and the remainder were prepared with the Illumina TruSeq PCR-free kit. All samples were sequenced on an Illumina NovaSeq 6000 instrument in 2x150 mode, to an approximate read depth of 50. After conversion to fastq, the sequenced reads were aligned to human reference genome GRCh37 with the GEM-mapper (v3, Marco-Sola et al. 2012) and were sorted and converted to CRAM format with samtools version 1.9 (Li et al. 2009). No additional refinements, such as duplicate read marking, base-quality score recalibration, or indel realignment were performed.

@@

Models were trained for 50 epochs unless noted otherwise, using the Adam optimizer, a learning rate of $5e^{-5}$, and batch size of 512. Models were implemented in PyTorch (v2.10, Paszke et al 2019).
